\textbf{Generalized additive models to analyze non-linear trends in biomedical longitudinal data using R: Beyond repeated measures ANOVA and Linear Mixed Models}

### 2022-01-04 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/response_to_reviewers/Response_to_reviewers.Rmd

We would like to thank the reviewer for the careful and thorough analysis of this manuscript to _Statistics in Medicine_ and for the thoughtful comments and constructive suggestions, which helped improve the quality of this manuscript. We carefully considered the reviewer’s comments and in this document, explain how we revised the manuscript based on those comments and suggestions.

# General Comments to the Editor

Dr. Platt,

The authors thank you for your determination that our manuscript may be suitable for resubmission in _Statistics in Medicine_ after addressing the reviewer's comments. To this end, we have addressed all critiques. We hope these revisions, submitted on January 4th, 2022, improve the manuscript so it is deemed worthy of publication in _Statistics in Medicine_. Following are our detailed responses to reviewer comments.

@@

L57 _p_ values cannot be significant or otherwise, they just are. They might indicate statistically significant effects, where the weight of evidence against the null hypothesis is sufficient to meet some threshold of “significance”. I note the use of scare quotes so perhaps this sentence was intended in jest but even
so it is confusing and unnecessary.

## \textcolor{reviewersblue} {Reply to Reviewer's Comments}

We appreciate the reviewer's comments here. However, the indicated sentence is not intended in jest. We are using in this sentence (and in the paper in general) a statistical approach that is understandable by biomedical researchers in general. To this day, the field values "significance" (defined by the arbitrarily set threshold of _p_<0.05). We are trying to convince the reader that low _p-values_ are possible with ill-fitted models, and we use language that is understandable by the reader.

The updated sentence in L59 now reads "significant effects (_p-value_<0.05)".

@@

## Reviewer's Comments

L293 You never really describe what “data-drive flexibility” is. Why are GAMs data driven and LMMs and GLMs not? This is a critical concept to convey I feel.

## \textcolor{reviewersblue} {Reply to Reviewer's Comments}

This line has been removed from the revised manuscript. However, we acknowledge that we didn't clearly conveyed this point to the read. In Section 6 (Discussion), L549-550 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided _a priori_ by the user.

### 2022-01-04 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/response_to_reviewers/Response_to_reviewers.tex

We would like to thank the reviewer for the careful and thorough analysis of this manuscript to \emph{Statistics in Medicine} and for the thoughtful comments and constructive suggestions, which helped improve the quality of this manuscript. We carefully considered the reviewer's comments and in this document, explain how we revised the manuscript based on those comments and suggestions.

\hypertarget{general-comments-to-the-editor}{%
\section{General Comments to the Editor}\label{general-comments-to-the-editor}}

Dr.~Platt,

The authors thank you for your determination that our manuscript may be suitable for resubmission in \emph{Statistics in Medicine} after addressing the reviewer's comments. To this end, we have addressed all critiques. We hope these revisions, submitted on January 4th, 2022, improve the manuscript so it is deemed worthy of publication in \emph{Statistics in Medicine}. Following are our detailed responses to reviewer comments.

@@

L57 \emph{p} values cannot be significant or otherwise, they just are. They might indicate statistically significant effects, where the weight of evidence against the null hypothesis is sufficient to meet some threshold of ``significance''. I note the use of scare quotes so perhaps this sentence was intended in jest but even
so it is confusing and unnecessary.

\hypertarget{section-18}{%
\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-18}}

We appreciate the reviewer's comments here. However, the indicated sentence is not intended in jest. We are using in this sentence (and in the paper in general) a statistical approach that is understandable by biomedical researchers in general. To this day, the field values ``significance'' (defined by the arbitrarily set threshold of \emph{p}\textless0.05). We are trying to convince the reader that low \emph{p-values} are possible with ill-fitted models, and we use language that is understandable by the reader.

The updated sentence in L59 now reads ``significant effects (\emph{p-value}\textless0.05)''.

@@

\hypertarget{reviewers-comments-23}{%
\subsection{Reviewer's Comments}\label{reviewers-comments-23}}

L293 You never really describe what ``data-drive flexibility'' is. Why are GAMs data driven and LMMs and GLMs not? This is a critical concept to convey I feel.

\hypertarget{section-24}{%
\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-24}}

This line has been removed from the revised manuscript. However, we acknowledge that we didn't clearly conveyed this point to the read. In Section 6 (Discussion), L549-550 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided \emph{a priori} by the user.

### 2022-01-04 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/Full_document_SIM_No_Appendix.tex
Updated

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\newcommand{\passthrough}[1]{#1}
\lstset{defaultdialect=[5.3]Lua}
\lstset{defaultdialect=[x86masm]Assembler}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{mathptmx}
\usepackage{lineno}
\usepackage[affil-it,blocks]{authblk}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{setspace}
%\usepackage[nomarkers,figuresonly]{endfloat}
%set a box to put the ORCID logo
\newbox{\myorcidaffilbox}
\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{latex_docs/orcid}}

@@

\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}}
\date{\vspace{-2.5em}}

@@

In contrast, generalized additive models (GAMs) relax the assumptions made by LMEMs and rm-ANOVA, and allow the data to determine the fit of the model while permitting incomplete observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze longitudinal data with non-linear trends in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We present the basic theory of GAMs and using reported trends of oxygen saturation in tumors, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R. We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when incomplete observations exist (with 40\% of the simulated observations missing). To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.

@@

A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from data with non-linear trends, the validity of such a metric is dependent on how adequate the model fits the data. In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}}. For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid\textsuperscript{\protect\hyperlink{ref-wang2019}{18}}.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Generalized additive models to analyze non-linear trends in biomedical longitudinal data using R: Beyond repeated measures ANOVA and Linear Mixed Models},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\newcommand{\passthrough}[1]{#1}
\lstset{defaultdialect=[5.3]Lua}
\lstset{defaultdialect=[x86masm]Assembler}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{mathptmx}
\usepackage{lineno}
\usepackage[affil-it,blocks]{authblk}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{setspace}
%\usepackage[nomarkers,figuresonly]{endfloat}
%set a box to put the ORCID logo
\newbox{\myorcidaffilbox}
\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{latex_docs/orcid}}

@@

\title{\textbf{Generalized additive models to analyze non-linear trends in biomedical longitudinal data using R: Beyond repeated measures ANOVA and Linear Mixed Models}}
\date{\vspace{-2.5em}}

@@

In contrast, generalized additive models (GAMs) relax the assumptions made by LMEMs and rm-ANOVA, and allow the data to determine the fit of the model while permitting incomplete observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze longitudinal data with non-linear trends in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on data with non-linear trends. We present the basic theory of GAMs and using reported trends of oxygen saturation in tumors, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R. We also show that GAMs are able to produce estimates with non-linear trends even when incomplete observations exist (with 40\% of the simulated observations missing). To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.

@@

A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant effects ( \emph{p-value}\textless0.05)'' from data with non-linear trends, the validity of such a metric is dependent on how adequate the model fits the data. In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}}. For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid\textsuperscript{\protect\hyperlink{ref-wang2019}{18}}.
